Hierarchical Region-Network Sparsity for\\ High-Dimensional Inference in Brain Imaging

### 2016-11-23 paper/ipmi2017/paper.tex
some news

\cite{jenatton2012multi} to create a set of convex sparse penalty terms.

@@

Major brain networks already emerge in human individuals around birth
\cite{doria2010}.
In adults,
the regions that compose
a given cohesive network have more
similar functional profiles than nodes from different networks
\cite{anderson2013}.
As a well-established method for network estimation,
ICA \cite{beckmann2005} yields continuous brain maps with
weights for each voxel. The region nodes of ICA
networks are spatially disjoint sets of voxel groups that
agree with boundaries of brain atlases.
Hence,
each region from a brain atlas can be uniquely associated with one of
the extracted ICA networks.
%
Here, previously published network definitions
obtained using ICA \cite{smith2009}
and
region definitions obtained from
spatially constrained clustering \cite{crad12}
allow constructing a hierarchy of global ICA networks with their
assigned local cluster regions
(Figure \ref{fig_priors}).
The ensuing network-region tree is used as a structural prior
of expected weight distributions
that guide supervised model fitting.

This tree structure can be plugged into
hierarchical sparsity penalty terms
\cite{jenatton2012multi}.
It extends the group lasso
\cite{yuan2006model}
by permitting variable groups that contain each other
in a nested tree structure.
The first hierarchical level are
the network groups with all
the voxels of the brain regions associated with them.
Each network node in turn descends into a second hierarchical level with
brain regions of neighboring voxels.
%
As a consequence of this region-network sparsity tree,
a child node enters the set of relevant voxel variables only
if its parent node has been selected
\cite{bach2012optimization}.
Conversely,
if a parent node is deselected,
also the voxel variables of all child nodes are deselected.
%
Moreover,
the coefficient of all region or all network groups can
be weighed individually.
Trading off the voxel penalties of the network level against the
voxel penalties of the region level,
we can design distinct estimation regimes.
%
% Setting a low penalty on the network groups makes it is probable
% that all of them are active in the estimated weight maps. If we then select
% higher penalties on region groups, selection of relevant region groups is
% forced without obeying network definitions.
% Conversely, setting low penalties on the region definitions
% makes it possible for
% all voxels to be active. Selecting higher penalties on the networks then
% leads to a selection of networks with all regions associated with it.
%
This region-network equilibrium allows inspecting the relative contributions
of the region and network levels
of the hierarchical tree prior.

% \paragraph{Problem formulation.}
\subsection{Problem formulation}
We formulate our estimation problem in the framework of regularized
empirical risk
minimization applied to linear models.
The goal is to estimate a good predictor of psychological tasks
given a single brain image. Let the set \(\mathcal X\in\R^{n \times p}\) represent brain
images of \(p > 0\) voxels.
%
We then minimize the risk \(\mathcal L(\hat y, y)\) with
\(\hat y = f(X\hat w + \hat b)\), where \(f\) is a link function 
(e.g., sigmoid for logistic regression, identity for linear regression),
and \(\mathcal L\) usually represents an appropriate negative loglikelihood.
We incorporate an informative prior through regularization:
\[\argmin_{w, b} \mathcal L(f(Xw + b), y) + \lambda\Omega(w),\]
where \(\lambda > 0\) and \(\Omega\) is the regularizer.

Brain \textit{regions} are defined as disjoint groups of voxels. Let \(\mathcal G\)
be a partition of \(\{1, \dots, p\}\), i.e.
\[  \bigcup_{i} g_i = \{1, \dots, p\} \textrm{ and } g_i\cap g_j=\emptyset
\quad\forall i\not=j\]

Brain \textit{networks} consist of one or several brain regions.
The set of brain networks \(\mathcal{H}\) also forms a partition of 
\(\{1, \dots, p\}\) that is consistent with \(\mathcal G\) in
the sense that
\[\forall g\in\mathcal G, h\in\mathcal H,\quad
\textrm{ either } g\subset h\textrm{ or }g\cap h = \emptyset.\]
This allows for an unambiguous assignment of each region \(g\in\mathcal G\) to one
network \(h\in\mathcal H\) and thus generates a tree structure.
A root node is added to contain all voxels.

For a brain image \(w\in\R^p\) and a group \(g\), the vector 
\(w_g\in\R^{|g|}\) is defined as the restriction of \(w\) to the coordinates
 in \(g\). The penalty structured by network
and region information can then be written as
\[\Omega(w) = \alpha\sum_{h\in\mathcal H}\eta_h\|w_h\|_2 + \beta\sum_{g\in\mathcal G}\eta_g\|w_g\|_2.\]
As originally recommended \cite{yuan2006model},
we set \(\eta_g = 1/\sqrt{|g|}\) to
account for discrepancy in group sizes. The hierarchy-level-specific factors 
\(\alpha > 0\) and \(\beta > 0\) can tradeoff region-weighted and
network-weighted models against each other.

Decreasing \(\alpha\) leads to less 
penalization of brain networks and thus the tendency for fully active groups
and dense brain maps. If at the same time \(\beta\) is increased to 
induce group sparsity, then only the structure of brain regions encoded
by \(\mathcal G\) is acknowledged. Conversely, if \(\beta\) is chosen 
sufficiently small and \(\alpha\) increased,
the detected structure will derive from \(\mathcal H\), leading to the
selection of brain networks rather than regions.

@@

%
\begin{figure*}
\vspace{-0.6cm}
\begin{centering}
% \includegraphics[width=1.00\textwidth]{../figures/sparsities.pdf}
\includegraphics[width=0.60\textwidth]{../figures/ROC_ALL_perclass.png}
\vspace{-0.6cm}
\caption{\textbf{Prediction performance across sparsity priors.}
Comparing the performance of logistic regression estimators
with 6 different structured and unstructered
sparse regularization penalties (\textit{colors})
in classifying neural activity from 18 psychological tasks.
The area under the curve (AUC)
is provided on an identical test set
as class-wise measure (\textit{y-axis}) and across-class mean (\textit{legend}).
%
% \textit{Hierarchical Tree Sparsity}: Structured
% $\ell_1/\ell_2$-block norm with a hierarchy of
% region and network priors
% exhibited the best out-of-sample performance
% ($\alpha = 1, \beta = 1$).
% \textit{Lasso}: Unstructured $\ell_1$-penalized logistic regression
% imposed a minimum of relevant brain voxels without
% assuming special structure.
% \textit{Elastic-Net}: Unstructured logistic regression
% with interpolation between $\ell_1$- and $\ell_2$-norm 
% imposed an equilibrium between sparsity and model fit.
% \textit{(Sparse) Group Sparsity}: Structured $\ell_1/\ell_2$-block
% norm (with additional $\ell_1$ term)
% imposed region compartments, but naive to network structure.
% \textit{Trace-norm}: Structured trace-norm penalization
% imposed low-rank structure
% with sparsity of unknown ``network'' patterns,
% but naive to region structure.
%
A-priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting
various psychological tasks from brain maps.
}
\end{centering}
\label{fig_sparsities}
\vspace{-0.6cm}
\end{figure*}
%
% \paragraph{Implementation.}
\subsection{Implementation}
All experiments were performed in Python.
We used \textit{nilearn} to process and resphape
the extensive neuroimaging data 
\cite{abrah14},
\textit{scikit-learn} to design machine-learning
data processing pipelines
\cite{pedr11},
and
\textit{SPAMs} for numerically optimized
implementations of the sparse learners
(\url{http://spams-devel.gforge.inria.fr/}).
All Python scripts that generated the results are
accessible online for reproducibility and reuse
(\url{http://github.com/anonymous/anonymous}).
% (\url{http://github.com/banilo/nips2015}).
\nopagebreak
\subsection{Data}
% \paragraph{Data.}
As the currently biggest open-access dataset in brain imagingf5,
we chose brain data from the HCP
\cite{barch2013}.
Neuroimaging data with labels of ongoing psychological processes
were drawn from 500
healthy HCP participants.
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants
(see appendix for detailed descriptions of dataset and tasks).
% across participants (Table \ref{table_tasks}).
The HCP data incorporated $n$ = 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were warped into a common $60\times72\times60$ space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
Hence,
the present experiments were based on task-labeled HCP maps of neural
activity with
$p$ = 79,941 voxels of interest that represent z-scored-values in gray matter.

>>>>

to create a set of convex sparse penalty terms.

@@

Major brain networks emerge in human individuals around birth
\cite{doria2010}.
Their nodes have more
similar functional profiles than nodes from different networks
\cite{anderson2013}.
As a popular method for network extraction,
ICA \cite{beckmann2005} yields continuous brain maps with
weights for each voxel. The region nodes of ICA
network are spatially disjoint sets of voxel groups that
agree with boundaries of brain atlases.
Hence,
each region from a brain atlas can be uniquely associated with one of
the extracted ICA networks.
%
Here, previously published network definitions
obtained using ICA \cite{smith2009}
and
region definitions obtained from
spatially constrained clustering \cite{crad12}
allowed constructing a hierarchy of global ICA networks with their
assigned local cluster regions
(Figure \ref{fig_priors}).
The ensuing network-region tree was used as a frequentist prior
of expected weight distributions
to advantageously bias supervised model fitting.

Specifically, this tree structure was plugged into
hierarchical sparsity penalty terms
\cite{jenatton2012multi}.
It extends the group lasso
\cite{yuan2006model}
by permitting variable groups that contain each other
in a nested tree structure.
The first hierarchical level are
the network groups with all
the voxels of the brain regions associated with them.
Each network node in turn descends into a second hierarchical level with
brain regions of neighboring voxels.
%
Induced by the region-network sparsity tree,
a child node enters the set of relevant voxel variables only
if its parent node has been selected
\cite{bach2012optimization}.
Conversely,
if a parent node is deselected,
also the voxel variables of all child nodes are deselected.
%
Moreover,
the coefficients of all region or all network groups can
be weighed individually.
Trading off the voxel penalties of the network level against the
voxel penalties of the region level
we can design distinct estimation regimes.
%
% Setting a low penalty on the network groups makes it is probable
% that all of them are active in the estimated weight maps. If we then select
% higher penalties on region groups, selection of relevant region groups is
% forced without obeying network definitions.
% Conversely, setting low penalties on the region definitions
% makes it possible for
% all voxels to be active. Selecting higher penalties on the networks then
% leads to a selection of networks with all regions associated with it.
%
% This region-network equilibrium allows inspecting the relative contributions
% of the region and network levels
% of the hierarchical tree prior.

% \paragraph{Problem formulation.}
\subsection{Problem formulation}
We formulate our estimation problem in the framework of regularized
empirical risk
minimization applied to linear models.
The goal is to estimate a good predictor of psychological tasks
given a single brain image. Let the set \(\mathcal X\in\R^{n \times p}\) represent brain
images of \(p > 0\) voxels.
We then minimize the risk \(\mathcal L(\hat y, y)\) with
\(\hat y = f(X\hat w + \hat b)\), where \(f\) is a link function 
(e.g., sigmoid for logistic regression, identity for linear regression),
and \(\mathcal L\) usually represents an appropriate negative loglikelihood.
We incorporate an informative prior through regularization:
\[\argmin_{w, b} \mathcal L(f(Xw + b), y) + \lambda\Omega(w),\]
where \(\lambda > 0\) and \(\Omega\) is the regularizer.
Brain \textit{regions} are defined as disjoint groups of voxels. Let \(\mathcal G\)
be a partition of \(\{1, \dots, p\}\), i.e.
\[  \bigcup_{i} g_i = \{1, \dots, p\} \textrm{ and } g_i\cap g_j=\emptyset
\quad\forall i\not=j\]
Brain \textit{networks} consist of one or several brain regions.
The set of brain networks \(\mathcal{H}\) also forms a partition of 
\(\{1, \dots, p\}\) that is consistent with \(\mathcal G\) in
the sense that
\[\forall g\in\mathcal G, h\in\mathcal H,\quad
\textrm{ either } g\subset h\textrm{ or }g\cap h = \emptyset.\]
This allows for an unambiguous assignment of each region \(g\in\mathcal G\) to one
network \(h\in\mathcal H\) and thus generates a tree structure.
A root node is added to contain all voxels.

For a brain image \(w\in\R^p\) and a group \(g\), the vector 
\(w_g\in\R^{|g|}\) is defined as the restriction of \(w\) to the coordinates
 in \(g\). The penalty structured by network
and region information can then be written as
\[\Omega(w) = \alpha\sum_{h\in\mathcal H}\eta_h\|w_h\|_2 + \beta\sum_{g\in\mathcal G}\eta_g\|w_g\|_2.\]
As originally recommended \cite{yuan2006model},
we set \(\eta_g = 1/\sqrt{|g|}\) to
account for discrepancy in group sizes. The hierarchy-level-specific factors 
\(\alpha > 0\) and \(\beta > 0\) can tradeoff region-weighted and
network-weighted models against each other.
Decreasing \(\alpha\) leads to less 
penalization of brain networks and thus the tendency for fully active groups
and dense brain maps. If at the same time \(\beta\) is increased to 
induce group sparsity, then only the structure of brain regions encoded
by \(\mathcal G\) is acknowledged. Conversely, if \(\beta\) is chosen 
sufficiently small and \(\alpha\) increased,
the detected structure will derive from \(\mathcal H\), leading to the
selection of brain networks rather than regions.

@@

%
\begin{figure*}
\vspace{-0.6cm}
\begin{centering}
% \includegraphics[width=1.00\textwidth]{../figures/sparsities.pdf}
\includegraphics[width=0.60\textwidth]{../figures/ROC_ALL_perclass.png}
\vspace{-0.6cm}
\caption{\textbf{Prediction performance across sparsity priors.}
Comparing the performance of logistic regression estimators
with 6 different structured and unstructered
sparse regularization penalties (\textit{colors})
in classifying neural activity from 18 psychological tasks.
The area under the curve (AUC)
is provided on an identical test set
as class-wise measure (\textit{y-axis}) and across-class mean (\textit{legend}).
%
% \textit{Hierarchical Tree Sparsity}: Structured
% $\ell_1/\ell_2$-block norm with a hierarchy of
% region and network priors
% exhibited the best out-of-sample performance
% ($\alpha = 1, \beta = 1$).
% \textit{Lasso}: Unstructured $\ell_1$-penalized logistic regression
% imposed a minimum of relevant brain voxels without
% assuming special structure.
% \textit{Elastic-Net}: Unstructured logistic regression
% with interpolation between $\ell_1$- and $\ell_2$-norm 
% imposed an equilibrium between sparsity and model fit.
% \textit{(Sparse) Group Sparsity}: Structured $\ell_1/\ell_2$-block
% norm (with additional $\ell_1$ term)
% imposed region compartments, but naive to network structure.
% \textit{Trace-norm}: Structured trace-norm penalization
% imposed low-rank structure
% with sparsity of unknown ``network'' patterns,
% but naive to region structure.
%
Simultaneous knowledge of both region and network neighborhoods
was hence most beneficial for predicting
tasks from neural activity.
}
\end{centering}
\label{fig_sparsities}
\vspace{-0.6cm}
\end{figure*}
%
% \paragraph{Implementation.}
\subsection{Implementation}
All experiments were performed in Python.
We used \textit{nilearn} to process and resphape
the extensive neuroimaging data 
\cite{abrah14},
\textit{scikit-learn} to design machine-learning
data processing pipelines
\cite{pedr11},
and
\textit{SPAMs} for numerically optimized
implementations of the sparse learners
(\url{http://spams-devel.gforge.inria.fr/}).
All Python scripts that generated the results are
accessible online for reproducibility and reuse
(\url{http://github.com/anonymous/anonymous}).
% (\url{http://github.com/banilo/nips2015}).
\nopagebreak
\subsection{Data}
% \paragraph{Data.}
As the currently biggest open-access dataset in brain imagingf5,
we chose brain data from the HCP
\cite{barch2013}.
Neuroimaging data with labels of ongoing psychological processes
were drawn from 500
healthy HCP participants.
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants
(see appendix for detailed descriptions of dataset and tasks).
% across participants (Table \ref{table_tasks}).
The HCP data incorporated $n$ = 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were warped into a common $60\times72\times60$ space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
Hence,
the present experiments were based on task-labeled HCP maps of neural
activity with
$p$ = 79,941 voxels of interest that represent z-scored-values in gray matter.
